Biogeography of crop progenitors and wild plant resources in Late Pleistocene--Early Holocene West Asia, 21--8.3 ka

### 2025-01-04 analysis/paper.tex
Improve PDF formatting

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  review]{elsarticle}

@@

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources in West Asia, 21--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources in West
Asia, 21--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

  \begin{itemize}
  \tightlist
  \item[$\square$]
    tbl-occ-count
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-climate-periods
  \end{itemize}
\item[$\square$]
  Clean up package code
\item[$\square$]
  Appendix with all hindcast predictions
\item[$\square$]
  References \& copyediting
\item[$\square$]
  Final proofread
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    But these have a variety of biases (human selection, taphonomy,
    etc.)
  \item
    And at the end of the day only represent specific places --
    interpolation to the entire region is not straightforward
  \item
    Overlap makes it difficult to see where human choices depart from
    the environmental background (cf. Martin et al., 2016)
  \item
    Or to fully contextualise subsistence strategies and shifts in
    strategies in response to environmental shifts (e.g. Yaworsky et
    al., 2023).
  \end{itemize}
\item
  Here we present an alternative approach using ENM

@@

  \begin{itemize}
  \tightlist
  \item
    Distinguish environmental from potentially anthropogenic change
    (\textbf{MartinEtAl2017?}; \textbf{MartinEtAl2025?})
  \item
    Reconstruct sequences of domestication (\textbf{YeomansEtAl2017?})
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

@@

Ecological niche modelling (ENM) or species distribution modelling (SDM)
is widely used by ecologists to predict the geographic range of a
species based on a set of environmental predictors. Essentially, it
involves combining records of where an organism has been observed with
environmental data (climate, topography, etc.) for those locations to
model the range of environmental values at which that species -- its
environmental niche. This model can then be used to predict the range of
the organism in question either in the same or a different environment.
(\textbf{CITE?}) suggests reserving the term `species distribution
modelling' for when the method is used to recover the verifiable range
of a species in a real and existing environment, and using `ecological
niche modelling' as the broader term covering hypothetical or predictive
applications -- a convention we follow here when referring to predictive
or `hindcast' models of past ranges. Within this overarching framework,
ecological niche modelling encompasses a wide range of applications and
a variety of potential environmental predictors, modelling approaches,
and methodologies, which we will not attempt to review here.

Ecological niche modelling has long been of interest to archaeologists
as both a means of exploring the biological niche of humans and for
reconstructing the past environments they inhabited (Franklin et al.,
2015; \textbf{DavidPollyEronen2011?}). In the first sense, it has been
used most extensively to model the range of humans and other hominin
species (e.g. \textbf{BenitoEtAl2017?}; \textbf{YousefiEtAl2020?};
\textbf{BanksEtAl2021?}; \textbf{YaworskyEtAl2024a?};
\textbf{YaworskyEtAl2024b?}; \textbf{GuranEtAl2024?}), especially in the
Palaeolithic. This overlaps with what archaeologists usually call
generically `predictive modelling'
(\textbf{VerhagenWhitley2020?})---more precisely `site distribution
modelling'---which is essentially the same approach as (and often
borrows methodologies from) ecological niche modelling but applied to
the occurrence of archaeological sites. Here what is modelled is not
strictly a biological niche alone, but also aspects of human geography,
taphonomy, and archaeological visibility. These applications can be
distinguihed from `palaeoecological niche modelling', where the object
of model remains, as in ecology, a non-human biological niche.

(Franklin et al., 2015) review palaeoecological niche modelling and
advocate for its greater adoption in environmental archaeology. One
relevant early example is (Conolly et al., 2012) used the occurrence of
wild and domestic \emph{Bos} remains at prehistoric archaeological sites
in Europe and West Asia to map the evolving niche of cattle over the
Pleistocene--Holocene transition. It has been used to model the
availability of fauna exploited by humans at wider scales (e.g.
\textbf{deAndresHerreroEtAl2018?}; \textbf{YaworksyEtAl2023?}) and, in a
West Asian context, of foraged plant resources in the landscape around
the Neolithic site of XX (Collins et al., 2018). Modelling the spread of
crops has been another significant archaeological application
(\textbf{CremaEtAl?}).

@@

The alternative approach is to train the model using contemporary
occurrence and environmental data and then use palaeoenvironmental data
to `hindcast' its predictions backwards in time. Like (Franklin et al.,
2015), we view the hindcasting approach as more promising, because
training datasets for both occurrences and environment are far more
readily available, complete and reliable for the present than the past.
There is some scepticism in the ecological niche modelling literature
about the ability of such models to make accurate predictions in unknown
environments (like the past) (\textbf{CITES?}), but here the hindcasting
approach also presents an opportunity: it reserves archaeological
occurrence data as an independent dataset that can be used to assess the
retrodictive performance of the model. This possibily was suggested by
(Franklin et al., 2015) but to our knowledge our study represents the
first attempt to actually do so.

The major practical limitation of the hindcasting approach is that it
relies on spatially explicit, high resolution palaeoenvironmental
surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most
applications, which is perhaps why only a minority of studies use it
(cf. Yaworsky et al., 2023). In this study, we are able to take
advantage of the increasing availability of high resolution, global
palaeoclimate data derived from simulation experiments with general
circulation models of climate (Brown et al., 2018;
\textbf{BrownEtAl2020?}; \textbf{KargerEtAl2023?}).

@@

@@

We consider X distinct taxa (Table~\ref{tbl-occ-count}) - all the
identifiable species known to be present at more than three Neolithic
sites in West Asia, according to our previous study (Arranz-Otaegui and
Roe, 2023).

Taxonomic names were resolves to the canonical form specified in the
GBIF Backbone Taxonomy (\textbf{GBIFSecretariat2023?}). So for example
occurrences for \emph{Bolboschoenus maritimus} also include those
recorded under the older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where known.

\subsection{Occurrence data}\label{occurrence-data-1}

\begin{verbatim}
Rows: 62 Columns: 5
-- Column specification --------------------------------------------------------
Delimiter: "\t"
chr (2): taxon, taxon_detail
dbl (3): n_present, p_present, n_occ

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

@@

Georeferenced occurrence data was obtained from the Global Biodiversity
Information Facility (GBIF) using via its application programming
interface and the R package `rgbif' (Chamberlain et al., 2024;
Chamberlain and Boettiger, 2017). GBIF occurrences marked has having
imprecise or duplicate coordinates were excluded from the training
dataset, as were fossil records. Although niche models have reasonable
predictive power even with small training samples (Hernandez et al.,
2006; Stockwell and Peterson, 2002; Wisz et al., 2008), we excluded 57
taxa with less than 50 occurrences in West Asia, following
recommendations for niche models generally and Random Forest-based
models specifically (Stockwell and Peterson, 2002;
\textbf{LuanEtAl2020?}). We also excluded one taxon (\emph{Avena
sterilis}) with over 50,000 occurrences, as this would have been
computationally prohibitive and we were uncertain what account for such
a disproportionately high number of records.

\begin{verbatim}
Rows: 35208 Columns: 11
-- Column specification --------------------------------------------------------
Delimiter: "\t"
chr (5): taxon, taxon_detail, scientific_name, genus, species
dbl (6): n_present, p_present, n_occ, gbif_key, longitude, latitude

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

Occurrence data only tells us where a species is present; there is
rarely definitive information on where the species is \emph{not} found.
We therefore need to generate random background points or
``pseudo-absences'' to feed to the model. There are several ways to do
this. We follow the advice of (Barbet-Massin et al., 2012) for
regression-based species distribution models and use a large (:10000)
random sample of points, weighted equally against the presences in the
regression. (Valavi et al., 2022) also recommend using a very large
background sample for random forest models.

@@

\begin{itemize}
\tightlist
\item
  Sixteen `bioclimatic' variables derived from monthly temperature and
  precipitation values, following standard practice for species
  distribution models (Hijmans et al., 2005). Contemporary bioclimatic
  predictor data for West Asia was extracted from the global CHELSA
  dataset (Karger et al., 2017), which predicts temperature and
  precipitation from downscaled general circulation model output at 1 km
  resolution.
\end{itemize}

\begin{itemize}
\tightlist
\item
  Terrain aspect and slope, which at high resolution perform well as
  proxies for solar radiation when modelling plant occurrence (Austin
  and Van Niel, 2011; Leempoel et al., 2015); and the topographic
  wetness index (TWI), which serves as a proxy for soil moisture and is
  particularly important in modelling arid environments (Campos et al.,
  2016; Di Virgilio et al., 2018; Kopecký and Čížková, 2010). All three
  were derived from the SRTM15+ digital elevation model using algorithms
  from WhiteboxTools (Lindsay, 2016).
\end{itemize}

\begin{itemize}
\tightlist
\item
  Edaphic data from SoilGrids (Hengl et al., 2017, 2014), which improves
  model performance for plants (Dubuis et al., 2013; Mod et al., 2016;
  Velazco et al., 2017). Based on a recent assessment of the reliability
  of SoilGrids data for species distribution modelling (Miller et al.,
  2024), we used a subset of four variables relating to soil texture
  (clay, silt, sand) and pH at the surface (0-5 cm depth).
\end{itemize}

Predictor data was transformed to the same projection system (WGS84 /
UTM 37 N) and a common resolution of X km.

For hindcasting, we used reconstructed bioclimatic data for 5 key
periods (\textbf{?@tbl-climate-periods}) generated from downscaled
paleoclimate simulations from the HadCM3 general circulation model
(Brown et al., 2018). Terrain and soil predictors were held constant,
since reconstructions of these variables in the past are not available
at sufficient scale. It is not likely that either macroscale topography
or soil characteristics have altered significantly over the period of
time considered here, so we assume that this does not degrade model
performance, and may in fact benefit it by providing `anchoring'
predictors that are independent of climate change.

@@

Ecological niche modelling is a classification problem that can be
approached with a wide range of statistical methods. A substantial
literature exists on the relatively performance of these approaches and
their respective parameterisations (reviewed in Valavi et al., 2022).
Random Forest, a widely-used machine learning algorithm, is amongst the
best performing approaches for presence-only species distribution
models, providing it is appropriately parameterised to account for the
class imbalance between presence and background samples (Valavi et al.,
2022, 2021). For our application, it also has the advantage of requiring
little to no manual parameter tuning to achieve good predictive results,
which makes it easier to model a larger numbers of taxa.

For each taxon we trained a classification model to predict occurrence
(presence or absence/background) based on our X predictor variables
(\textbf{?@tbl-predictors}). We used the Random Forest algorithm
implemented in the R package `ranger' (Wright and Ziegler, 2017) and the
`tidymodels' (\textbf{tidymodels?}) framework for data preprocessing and
model selection. To avoid overfitting, we follow (Valavi et al., 2021)
in their recommended hyperparameters and use of down-sampling to balance
presence and background samples. Models for each taxon were fit
independently, with redundant zero-variance predictors excluded, and
assessed based on balanced training (3/4) and test (1/4) partitions.

The output of the model is probabilistic. However, this should not be
understood as an actual probability of occurrence (\textbf{CITE?}), but
more akin to an estimate of habitat suitability. To simplify
interpretation, we can convert these predictions into binary
presence/absence maps, a process called ``thresholding''. We select the
threshold value for each model individually, using MaxSSS (as
recommended by Liu et al., 2013). This also makes it possible to
analyses the predictions together as an assemblage.

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-ArranzOtaeguiRoe2023}
Arranz-Otaegui, A., Roe, J., 2023. Revisiting the concept of the
{``{Neolithic Founder Crops}''} in southwest {Asia}. Vegetation History
and Archaeobotany. \url{https://doi.org/10.1007/s00334-023-00917-1}

\bibitem[\citeproctext]{ref-AustinVanNiel2011}
Austin, M.P., Van Niel, K.P., 2011. Improving species distribution
models for climate change studies: Variable selection and scale. Journal
of Biogeography 38, 1--8.
\url{https://doi.org/10.1111/j.1365-2699.2010.02416.x}

\bibitem[\citeproctext]{ref-BarbetMassinEtAl2012}
Barbet-Massin, M., Jiguet, F., Albert, C.H., Thuiller, W., 2012.
Selecting pseudo-absences for species distribution models: How, where
and how many?: {How} to use pseudo-absences in niche modelling? Methods
Ecol. Evol. 3, 327--338.
\url{https://doi.org/10.1111/j.2041-210X.2011.00172.x}

\bibitem[\citeproctext]{ref-BrownEtAl2018}
Brown, J.L., Hill, D.J., Dolan, A.M., Carnaval, A.C., Haywood, A.M.,
2018. {PaleoClim}, high spatial resolution paleoclimate surfaces for
global land areas. Sci Data 5, 180254.
\url{https://doi.org/10.1038/sdata.2018.254}

\bibitem[\citeproctext]{ref-CamposEtAl2016}
Campos, V.E., Cappa, F.M., Viviana, F.M., Giannoni, S.M., 2016. Using
remotely sensed data to model suitable habitats for tree species in a
desert environment. Journal of Vegetation Science 27, 200--210.
\url{https://doi.org/10.1111/jvs.12328}

\bibitem[\citeproctext]{ref-rgbif}
Chamberlain, S., Barve, V., Mcglinn, D., Oldoni, D., Desmet, P.,
Geffert, L., Ram, K., 2024.
\href{https://CRAN.R-project.org/package=rgbif}{Rgbif: Interface to the
global biodiversity information facility API}.

\bibitem[\citeproctext]{ref-ChamberlainBoettiger2017}
Chamberlain, S., Boettiger, C., 2017.
\href{https://doi.org/10.7287/peerj.preprints.3304v1}{R python, and ruby
clients for GBIF species occurrence data}. PeerJ PrePrints.

\bibitem[\citeproctext]{ref-CollinsEtAl2018}
Collins, C., Asouti, E., Grove, M., Kabukcu, C., Bradley, L.,
Chiverrell, R., 2018. Understanding resource choice at the transition
from foraging to farming: {An} application of palaeodistribution
modelling to the {Neolithic} of the {Konya Plain}, south-central
{Anatolia}, {Turkey}. J. Archaeol. Sci. 96, 57--72.
\url{https://doi.org/10.1016/j.jas.2018.02.003}

\bibitem[\citeproctext]{ref-ConollyEtAl2012}
Conolly, J., Manning, K., Colledge, S., Dobney, K., Shennan, S., 2012.
Species distribution modelling of ancient cattle from early {Neolithic}
sites in {SW Asia} and {Europe}. Holocene 22, 997--1010.
\url{https://doi.org/10.1177/0959683612437871}

\bibitem[\citeproctext]{ref-DiVirgilioEtAl2018}
Di Virgilio, G., Wardell-Johnson, G.W., Robinson, T.P., Temple-Smith,
D., Hesford, J., 2018. Characterising fine-scale variation in plant
species richness and endemism across topographically complex, semi-arid
landscapes. Journal of Arid Environments 156, 59--68.
\url{https://doi.org/10.1016/j.jaridenv.2018.04.005}

\bibitem[\citeproctext]{ref-DubuisEtAl2013}
Dubuis, A., Giovanettina, S., Pellissier, L., Pottier, J., Vittoz, P.,
Guisan, A., 2013. Improving the prediction of plant species distribution
and community composition by adding edaphic to topo-climatic variables.
Journal of Vegetation Science 24, 593--606.
\url{https://doi.org/10.1111/jvs.12002}

\bibitem[\citeproctext]{ref-FranklinEtAl2015}
Franklin, J., Potts, A.J., Fisher, E.C., Cowling, R.M., Marean, C.W.,
2015. Paleodistribution modeling in archaeology and paleoanthropology.
Quat. Sci. Rev. 110, 1--14.
\url{https://doi.org/10.1016/j.quascirev.2014.12.015}

\bibitem[\citeproctext]{ref-HenglEtAl2014}
Hengl, T., de Jesus, J.M., MacMillan, R.A., Batjes, N.H., Heuvelink,
G.B.M., Ribeiro, E., Samuel-Rosa, A., Kempen, B., Leenaars, J.G.B.,
Walsh, M.G., Gonzalez, M.R., 2014. {SoilGrids1km--global} soil
information based on automated mapping. PLoS One 9, e105992.
\url{https://doi.org/10.1371/journal.pone.0105992}

\bibitem[\citeproctext]{ref-HenglEtAl2017}
Hengl, T., Mendes de Jesus, J., Heuvelink, G.B.M., Ruiperez Gonzalez,
M., Kilibarda, M., Blagotić, A., Shangguan, W., Wright, M.N., Geng, X.,
Bauer-Marschallinger, B., Guevara, M.A., Vargas, R., MacMillan, R.A.,
Batjes, N.H., Leenaars, J.G.B., Ribeiro, E., Wheeler, I., Mantel, S.,
Kempen, B., 2017. {SoilGrids250m}: {Global} gridded soil information
based on machine learning. PLoS One 12, e0169748.
\url{https://doi.org/10.1371/journal.pone.0169748}

\bibitem[\citeproctext]{ref-HernandezEtAl2006}
Hernandez, P.A., Graham, C.H., Master, L.L., Albert, D.L., 2006. The
effect of sample size and species characteristics on performance of
different species distribution modeling methods. Ecography 29, 773--785.
\url{https://doi.org/10.1111/j.0906-7590.2006.04700.x}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  review]{elsarticle}

@@

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

@@

\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources in Late Pleistocene--Early Holocene West Asia, 21--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources in Late
Pleistocene--Early Holocene West Asia, 21--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

  \begin{itemize}
  \tightlist
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-climate-periods
  \end{itemize}
\item[$\square$]
  Clean up package code
\item[$\square$]
  Appendix with all hindcast predictions
\item[$\square$]
  References \& copyediting
\item[$\square$]
  Final proofread
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    But these have a variety of biases (human selection, taphonomy,
    etc.)
  \item
    And at the end of the day only represent specific places --
    interpolation to the entire region is not straightforward
  \item
    Overlap makes it difficult to see where human choices depart from
    the environmental background \citep[cf.][]{MartinEtAl2016}
  \item
    Or to fully contextualise subsistence strategies and shifts in
    strategies in response to environmental shifts
    \citep[e.g.][]{YaworskyEtAl2023}.
  \end{itemize}
\item
  Here we present an alternative approach using ENM

@@

  \begin{itemize}
  \tightlist
  \item
    Distinguish environmental from potentially anthropogenic change
    \citep{MartinEtAl2017, MartinEtAl2025}
  \item
    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

@@

Ecological niche modelling (ENM) or species distribution modelling (SDM)
is widely used by ecologists to predict the geographic range of a
species based on a set of environmental predictors. Essentially, it
involves combining records of where an organism has been observed with
environmental data (climate, topography, etc.) for those locations to
model the range of environmental values at which that species -- its
environmental niche. This model can then be used to predict the range of
the organism in question either in the same or a different environment.
\citet{CITE} suggests reserving the term `species distribution
modelling' for when the method is used to recover the verifiable range
of a species in a real and existing environment, and using `ecological
niche modelling' as the broader term covering hypothetical or predictive
applications -- a convention we follow here when referring to predictive
or `hindcast' models of past ranges. Within this overarching framework,
ecological niche modelling encompasses a wide range of applications and
a variety of potential environmental predictors, modelling approaches,
and methodologies, which we will not attempt to review here.

Ecological niche modelling has long been of interest to archaeologists
as both a means of exploring the biological niche of humans and for
reconstructing the past environments they inhabited
\citep{DavidPollyEronen2011, FranklinEtAl2015}. In the first sense, it
has been used most extensively to model the range of humans and other
hominin species
\citep[e.g.][]{BenitoEtAl2017, YousefiEtAl2020, BanksEtAl2021, YaworskyEtAl2024a, YaworskyEtAl2024b, GuranEtAl2024},
especially in the Palaeolithic. This overlaps with what archaeologists
usually call generically `predictive modelling'
\citep{VerhagenWhitley2020}---more precisely `site distribution
modelling'---which is essentially the same approach as (and often
borrows methodologies from) ecological niche modelling but applied to
the occurrence of archaeological sites. Here what is modelled is not
strictly a biological niche alone, but also aspects of human geography,
taphonomy, and archaeological visibility. These applications can be
distinguihed from `palaeoecological niche modelling', where the object
of model remains, as in ecology, a non-human biological niche.

\citet{FranklinEtAl2015} review palaeoecological niche modelling and
advocate for its greater adoption in environmental archaeology. In an
early application to West Asia, \citet{ConollyEtAl2012} used the
occurrence of wild and domestic \emph{Bos} remains at prehistoric
archaeological sites to map the evolving niche of cattle over the
Pleistocene--Holocene transition. It has been used to model the
availability of fauna exploited by humans at wider scales
\citep[e.g.][]{deAndresHerreroEtAl2018, YaworksyEtAl2023} and, in a West
Asian context, of foraged plant resources in the landscape around the
Neolithic site of XX \citep{CollinsEtAl2018}. Modelling the spread of
crops has been another significant archaeological application
\citep{CremaEtAl}.

@@

The alternative approach is to train the model using contemporary
occurrence and environmental data and then use palaeoenvironmental data
to `hindcast' its predictions backwards in time. Like
\citet{FranklinEtAl2015}, we view the hindcasting approach as more
promising, because training datasets for both occurrences and
environment are far more readily available, complete and reliable for
the present than the past. There is some scepticism in the ecological
niche modelling literature about the ability of such models to make
accurate predictions in unknown environments (like the past)
\citep{CITES}, but here the hindcasting approach also presents an
opportunity: it reserves archaeological occurrence data as an
independent dataset that can be used to assess the retrodictive
performance of the model. This possibily was suggested by
\citet{FranklinEtAl2015} but to our knowledge our study represents the
first attempt to actually do so.

The major practical limitation of the hindcasting approach is that it
relies on spatially explicit, high resolution palaeoenvironmental
surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most
applications, which is perhaps why only a minority of studies use it
\citep[cf.][]{YaworskyEtAl2023}. In this study, we are able to take
advantage of the increasing availability of high resolution, global
palaeoclimate data derived from simulation experiments with general
circulation models of climate
\citep{BrownEtAl2018, BrownEtAl2020, KargerEtAl2023}.

@@

We began by considering 62 distinct taxa (Table~\ref{tbl-occ-count}) -
all the identifiable species known to be present at more than three
Neolithic (c.~11.7--6.5 ka) sites in West Asia, according to our
previous study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. Taxonomic names were resolved to the
canonical form specified in the GBIF Backbone Taxonomy
\citep{GBIFSecretariat2023}. So for example occurrences for
\emph{Bolboschoenus maritimus} also include those recorded under the
older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where known.

@@

@@

Georeferenced occurrence data was obtained from the Global Biodiversity
Information Facility (GBIF) using via its application programming
interface and the R package `rgbif'
\citep{ChamberlainBoettiger2017, rgbif}. GBIF occurrences marked has
having imprecise or duplicate coordinates were excluded from the
training dataset, as were fossil records. Although niche models have
reasonable predictive power even with small training samples
\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
excluded 57 taxa with less than 50 occurrences in West Asia, following
recommendations for niche models generally and Random Forest-based
models specifically \citep{StockwellPeterson2002, LuanEtAl2020}. We also
excluded one taxon (\emph{Avena sterilis}) with over 50,000 occurrences,
as this would have been computationally prohibitive and we were
uncertain what account for such a disproportionately high number of
records.

Occurrence data only tells us where a species is present; there is
rarely definitive information on where the species is \emph{not} found.
We therefore need to generate random background points or
``pseudo-absences'' to feed to the model. There are several ways to do
this. We follow the advice of \citet{BarbetMassinEtAl2012} for
regression-based species distribution models and use a large (:10000)
random sample of points, weighted equally against the presences in the
regression. \citet{ValaviEtAl2022} also recommend using a very large
background sample for random forest models.

@@

\begin{itemize}
\tightlist
\item
  Sixteen `bioclimatic' variables derived from monthly temperature and
  precipitation values, following standard practice for species
  distribution models \citep{HijmansEtAl2005}. Contemporary bioclimatic
  predictor data for West Asia was extracted from the global CHELSA
  dataset \citep{KargerEtAl2017}, which predicts temperature and
  precipitation from downscaled general circulation model output at 1 km
  resolution.
\end{itemize}

\begin{itemize}
\tightlist
\item
  Terrain aspect and slope, which at high resolution perform well as
  proxies for solar radiation when modelling plant occurrence
  \citep{AustinVanNiel2011, LeempoelEtAl2015}; and the topographic
  wetness index (TWI), which serves as a proxy for soil moisture and is
  particularly important in modelling arid environments
  \citep{KopeckyCizkova2010, CamposEtAl2016, DiVirgilioEtAl2018}. All
  three were derived from the SRTM15+ digital elevation model using
  algorithms from WhiteboxTools \citep{Lindsay2016}.
\end{itemize}

\begin{itemize}
\tightlist
\item
  Edaphic data from SoilGrids \citep{HenglEtAl2014, HenglEtAl2017},
  which improves model performance for plants
  \citep{DubuisEtAl2013, ModEtAl2016, VelazcoEtAl2017}. Based on a
  recent assessment of the reliability of SoilGrids data for species
  distribution modelling \citep{MillerEtAl2024}, we used a subset of
  four variables relating to soil texture (clay, silt, sand) and pH at
  the surface (0-5 cm depth).
\end{itemize}

Predictor data was transformed to common equal-area projection and
resolution of 5 km.

\caption{\label{tbl-climate-periods}}

\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr}
\toprule
Period & Age, ka \\ 
\midrule\addlinespace[2.5pt]
Last Glacial Maximum (LGM) & 21.0–17.0 \\ 
Heinrich Stadial 1 (HS1) & 17.0–14.7 \\ 
Bølling-Allerød (BA) & 14.7–12.9 \\ 
Younger Dryas (YDS) & 12.9–11.7 \\ 
Early Holocene (EH) & 11.7–8.3 \\ 
Current (CUR) & — \\ 
\bottomrule
\end{tabular*}

Palaeoclimatic periods used for hindcasting, after \citet{BrownEtAl2018}

For hindcasting, we used reconstructed bioclimatic data for 5 key
periods (Table~\ref{tbl-climate-periods}) generated from downscaled
paleoclimate simulations from the HadCM3 general circulation model
\citep{BrownEtAl2018}. Terrain and soil predictors were held constant,
since reconstructions of these variables in the past are not available
at sufficient scale. It is not likely that either macroscale topography
or soil characteristics have altered significantly over the period of
time considered here, so we assume that this does not degrade model
performance, and may in fact benefit it by providing `anchoring'
predictors that are independent of climate change.

@@

Ecological niche modelling is a classification problem that can be
approached with a wide range of statistical methods. A substantial
literature exists on the relatively performance of these approaches and
their respective parameterisations \citep[reviewed
in][]{ValaviEtAl2022}. Random Forest, a widely-used machine learning
algorithm, is amongst the best performing approaches for presence-only
species distribution models, providing it is appropriately parameterised
to account for the class imbalance between presence and background
samples \citep{ValaviEtAl2021, ValaviEtAl2022}. For our application, it
also has the advantage of requiring little to no manual parameter tuning
to achieve good predictive results, which makes it easier to model a
larger numbers of taxa.

For each taxon we trained a classification model to predict occurrence
(presence or absence/background) based on our X predictor variables
(\textbf{?@tbl-predictors}). We used the Random Forest algorithm
implemented in the R package `ranger' \citep{WrightZiegler2017} and the
`tidymodels' \citep{tidymodels} framework for data preprocessing and
model selection. To avoid overfitting, we follow \citet{ValaviEtAl2021}
in their recommended hyperparameters and use of down-sampling to balance
presence and background samples. Models for each taxon were fit
independently, with redundant zero-variance predictors excluded, and
assessed based on balanced training (¾) and test (¼) partitions.

The output of the model is probabilistic. However, this should not be
understood as an actual probability of occurrence \citep{CITE}, but more
akin to an estimate of habitat suitability. To simplify interpretation,
we can convert these predictions into binary presence/absence maps, a
process called ``thresholding''. We select the threshold value for each
model individually, using MaxSSS \citep[as recommended
by][]{LiuEtAl2013}. This also makes it possible to analyses the
predictions together as an assemblage.

@@

\renewcommand\refname{References}
  \bibliography{references.bib}
